---
title: "CNMI Child Coverage and Vaccination Planning"
author: "USAPI Assessment Team | Surveillance & Epidemiology Branch | ISD | NCIRD | CDC"
date: today
date-format: long
params:
  age_as_of_date: "2026-08-25"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    df-print: paged
    css: ../www/styles.css
execute:
  warning: false
  message: false
editor:
  markdown:
    wrap: 80
---

This Quarto document imports the three CNMI WebIZ reports, creates the
patient-level vaccination indicators, applies the CNMI geography rules,
exports the operational Excel files, and refreshes the dashboard data.

# Setup

```{r}
#| label: setup
#| include: false

required_packages <- c(
  "dplyr", "tidyr", "purrr", "stringr", "lubridate", "readxl",
  "readr", "janitor", "openxlsx", "here", "tibble", "scales"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Install the following packages before rendering:\n  install.packages(c(",
      paste0('"', missing_packages, '"', collapse = ", "),
      "))"
    ),
    call. = FALSE
  )
}

library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(lubridate)
library(readxl)
library(readr)
library(janitor)
library(openxlsx)
library(here)
library(tibble)

source(here::here("R", "cnmi_functions.R"), local = TRUE)

project_dir <- here::here()
data_dir <- here::here("data")
raw_data_dir <- here::here("data", "Raw Patient Data")
output_dir <- here::here("data", "Analytic Code Output")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

analysis_date <- as.Date(params$age_as_of_date)
if (is.na(analysis_date)) {
  stop("`params$age_as_of_date` must use YYYY-MM-DD format.", call. = FALSE)
}

date_stamp <- format(analysis_date, "%m%d%y")

neighborhood_file <- here::here("data", "CNMI_Neighborhood_Coordinates.xlsx")
region_file <- here::here("data", "CNMI_County_Population_Area_GIS.xlsx")

required_reference_files <- c(neighborhood_file, region_file)
missing_reference_files <- required_reference_files[!file.exists(required_reference_files)]

if (length(missing_reference_files) > 0) {
  stop(
    paste("Required geography file(s) missing:", paste(missing_reference_files, collapse = "\n")),
    call. = FALSE
  )
}
```

# Input files

The newest matching workbook is selected automatically. File names can include
a run date; they only need to begin with `Roster`, `Details` (or `Detail`), and
`RR` (or contain `Reminder`).

```{r}
#| label: input-files
#| include: false

roster_file <- newest_matching_file(raw_data_dir, "^Roster.*\\.xlsx$")
detail_file <- newest_matching_file(raw_data_dir, "^Details?.*\\.xlsx$")
reminder_file <- newest_matching_file(
  raw_data_dir,
  "(^RR.*|.*Reminder.*)\\.xlsx$"
)

input_inventory <- tibble(
  report = c("Patient roster", "Patient detail with services", "Reminder/recall"),
  file = basename(c(roster_file, detail_file, reminder_file))
)

roster_raw <- readxl::read_excel(roster_file, guess_max = 100000) |>
  janitor::clean_names()
detail_raw <- readxl::read_excel(detail_file, guess_max = 100000) |>
  janitor::clean_names()
reminder_raw <- readxl::read_excel(reminder_file, guess_max = 100000) |>
  janitor::clean_names()

if (!"dob" %in% names(roster_raw)) {
  stop("The roster file must contain a DOB column.", call. = FALSE)
}

neighborhood_lookup <- read_neighborhood_lookup(neighborhood_file)
```

```{r}
#| label: input-inventory
knitr::kable(input_inventory, caption = "Input reports selected for this run")
```

# Clean and link patient records

```{r}
#| label: clean-and-link
#| include: false

roster_clean <- roster_raw |>
  mutate(
    patient_id = extract_patient_id(roster_raw),
    dob = parse_cnmi_date(dob)
  )

if ("clinic_status" %in% names(roster_clean)) {
  roster_clean <- roster_clean |>
    filter(is.na(clinic_status) | str_to_upper(clinic_status) == "ACTIVE")
}

roster_clean <- roster_clean |>
  filter(!is.na(patient_id), !is.na(dob)) |>
  assign_cnmi_geography(neighborhood_lookup)

vaccine_history <- build_cnmi_vaccine_history(detail_raw)

reminder_recommendations <- coalesce_character_columns(
  reminder_raw,
  c(
    "immunization_recommendations", "immunization_recommendation",
    "recommendations", "recommendation"
  )
)

reminder_clean <- reminder_raw |>
  mutate(
    patient_id = extract_patient_id(reminder_raw),
    immunization_recommendations = reminder_recommendations
  ) |>
  filter(!is.na(patient_id)) |>
  group_by(patient_id) |>
  summarise(
    immunization_recommendations = paste(
      sort(unique(na.omit(immunization_recommendations))),
      collapse = "; "
    ),
    onreminder = 1L,
    .groups = "drop"
  )

analytic_full <- roster_clean |>
  left_join(vaccine_history, by = "patient_id") |>
  left_join(reminder_clean, by = "patient_id") |>
  add_cnmi_utd_flags(analysis_date) |>
  add_cnmi_need_and_product_flags(analysis_date) |>
  arrange(region, county, age_months, patient_id)

duplicate_patient_ids <- analytic_full |>
  count(patient_id, name = "rows") |>
  filter(rows > 1)

if (nrow(duplicate_patient_ids) > 0) {
  stop(
    paste0(
      "The roster contains duplicate patient IDs after cleaning. Review ",
      "duplicate_patient_ids before exporting."
    ),
    call. = FALSE
  )
}
```

# CNMI schedule rules used

The analysis applies the following CNMI-specific rules:

- MMR dose 1 is due beginning at 12 months; MMR dose 2 is not due before 48
  months.
- Varicella follows the same routine age anchors: dose 1 at 12-15 months and
  dose 2 at 4-6 years.
- HepA begins at 12 months. For a child with one recorded HepA dose, dose 2 is
  not marked due until at least six months after dose 1.

Source: [AAP Recommended Child and Adolescent Immunization Schedule,
2026](https://downloads.aap.org/AAP/PDF/AAP-Immunization-Schedule.pdf).

# Export patient-level files

```{r}
#| label: export-patient-files
#| include: false

full_dataset_file <- file.path(
  output_dir,
  paste0("Dataset 1_CNMI_2-83 Mos_Full Dataset_", date_stamp, ".xlsx")
)

deid_dataset_file <- file.path(
  output_dir,
  paste0("Dataset 2_CNMI_2-83 Mos_DeID_Full_Dataset_", date_stamp, ".xlsx")
)

patients_needing_file <- file.path(
  output_dir,
  paste0("CNMI_Patients_Needing_Vaccination_", date_stamp, ".xlsx")
)

missing_coordinates_file <- file.path(output_dir, "Missing_Coordinates.xlsx")

direct_identifier_columns <- c(
  "patient_name", "first_name", "middle_name", "last_name", "full_name",
  "primary_contact_lastname_firstname", "primary_contact_name",
  "telephone", "cell_phone", "phone", "email",
  "address_line_1", "address_line_2", "address", "zip", "postal_code",
  "medical_record_number", "mrn", "dob"
)

patient_id_crosswalk <- analytic_full |>
  distinct(patient_id) |>
  arrange(patient_id) |>
  mutate(dashboard_patient_id = sprintf("CNMI%06d", row_number()))

analytic_deid <- analytic_full |>
  left_join(patient_id_crosswalk, by = "patient_id") |>
  select(-any_of(c("patient_id", direct_identifier_columns))) |>
  rename(patient_id = dashboard_patient_id) |>
  relocate(patient_id)

need_columns <- unname(CNMI_VACCINE_NEED_MAP)
product_columns <- unname(CNMI_PRODUCT_MAP)

patients_needing <- analytic_full |>
  filter(
    onreminder == 1 |
      if_any(all_of(need_columns), ~dplyr::coalesce(as.logical(.x), FALSE))
  ) |>
  select(
    any_of(c(
      "patient_id", "first_name", "last_name", "patient_name", "dob",
      "age_months", "age_group_label", "region", "county",
      "primary_contact_lastname_firstname", "telephone", "cell_phone",
      "address_line_1", "address_line_2", "immunization_recommendations"
    )),
    all_of(need_columns),
    all_of(product_columns),
    any_of(c("mmr1utd", "mmr_utd", "var_utd", "hepa_utd"))
  ) |>
  # Do not need MMR2UTD in this operational file as in other jurisdictions.
  select(-any_of("mmr2utd"))

missing_coordinates <- analytic_deid |>
  filter(county != "Unknown", is.na(latitude) | is.na(longitude)) |>
  distinct(region, county) |>
  arrange(region, county)

openxlsx::write.xlsx(
  analytic_full, full_dataset_file,
  asTable = nrow(analytic_full) > 0, overwrite = TRUE
)
openxlsx::write.xlsx(
  analytic_deid, deid_dataset_file,
  asTable = nrow(analytic_deid) > 0, overwrite = TRUE
)
openxlsx::write.xlsx(
  patients_needing, patients_needing_file,
  asTable = nrow(patients_needing) > 0, overwrite = TRUE
)
openxlsx::write.xlsx(
  missing_coordinates, missing_coordinates_file,
  asTable = nrow(missing_coordinates) > 0, overwrite = TRUE
)
```

# Create dashboard data

```{r}
#| label: create-dashboard-data
#| include: false

source(here::here("R", "prepare_data.R"), local = TRUE)
```

# Create formatted Excel tables

```{r}
#| label: create-excel-tables
#| include: false

region_levels <- c("CNMI", "Saipan", "Rota", "Tinian")
age_levels <- c(
  "2-3 months", "4-5 months", "6-11 months", "12-18 months",
  "19-35 months", "3 years", "4 years", "5 years", "6 years"
)

dose_coverage_label <- function(age_group, vaccine) {
  case_when(
    age_group == "2-3 months" & vaccine == "DTaP" ~ "≥1 DTaP",
    age_group == "2-3 months" & vaccine == "IPV" ~ "≥1 IPV",
    age_group == "2-3 months" & vaccine == "Hib" ~ "≥1 Hib",
    age_group == "2-3 months" & vaccine == "HepB" ~ "≥2 HepB",
    age_group == "2-3 months" & vaccine == "PCV" ~ "≥1 PCV",
    age_group == "2-3 months" & vaccine == "Rotavirus" ~ "≥1 Rotavirus",

    age_group == "4-5 months" & vaccine == "DTaP" ~ "≥2 DTaP",
    age_group == "4-5 months" & vaccine == "IPV" ~ "≥2 IPV",
    age_group == "4-5 months" & vaccine == "Hib" ~ "≥2 Hib",
    age_group == "4-5 months" & vaccine == "HepB" ~ "≥2 HepB",
    age_group == "4-5 months" & vaccine == "PCV" ~ "≥2 PCV",
    age_group == "4-5 months" & vaccine == "Rotavirus" ~ "≥2 Rotavirus",

    age_group == "6-11 months" & vaccine == "DTaP" ~ "≥3 DTaP",
    age_group == "6-11 months" & vaccine == "IPV" ~ "≥3 IPV",
    age_group == "6-11 months" & vaccine == "Hib" ~ "≥2 Hib",
    age_group == "6-11 months" & vaccine == "HepB" ~ "≥3 HepB",
    age_group == "6-11 months" & vaccine == "PCV" ~ "≥3 PCV",
    age_group == "6-11 months" & vaccine == "Rotavirus" ~ "≥3 Rotavirus",

    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "DTaP" ~ "≥4 DTaP",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "IPV" ~ "≥3 IPV",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "MMR" ~ "≥1 MMR",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "Hib" ~ "≥3 Hib",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "HepB" ~ "≥3 HepB",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "PCV" ~ "≥4 PCV",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "Rotavirus" ~ "≥3 Rotavirus",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "Varicella" ~ "≥1 Varicella",
    age_group %in% c("12-18 months", "19-35 months", "3 years") & vaccine == "HepA" ~ "≥1 HepA (dose 2 if due)",

    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "DTaP" ~ "≥4 DTaP*",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "IPV" ~ "≥3 IPV*",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "MMR" ~ "≥2 MMR",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "Hib" ~ "≥3 Hib",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "HepB" ~ "≥3 HepB",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "PCV" ~ "≥4 PCV",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "Rotavirus" ~ "≥3 Rotavirus",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "Varicella" ~ "≥2 Varicella",
    age_group %in% c("4 years", "5 years", "6 years") & vaccine == "HepA" ~ "≥1 HepA (dose 2 if due)",
    TRUE ~ NA_character_
  )
}

with_cnmi_total <- function(data, grouping, summarise_expression) {
  region_rows <- data |>
    group_by(across(all_of(c("region", grouping)))) |>
    summarise(!!!summarise_expression, .groups = "drop")

  total_rows <- data |>
    group_by(across(all_of(grouping))) |>
    summarise(!!!summarise_expression, .groups = "drop") |>
    mutate(region = "CNMI", .before = 1)

  bind_rows(total_rows, region_rows) |>
    mutate(region = factor(region, levels = region_levels)) |>
    arrange(region, !!!rlang::syms(grouping)) |>
    mutate(region = as.character(region))
}

denominator_table <- with_cnmi_total(
  analytic_deid,
  "age_group_label",
  rlang::list2(Children = rlang::expr(n_distinct(patient_id)))
) |>
  mutate(age_group_label = factor(age_group_label, levels = age_levels)) |>
  arrange(factor(region, levels = region_levels), age_group_label) |>
  transmute(Region = region, `Age group` = as.character(age_group_label), Children)

reminder_by_village_detail <- analytic_deid |>
  filter(onreminder == 1) |>
  count(region, county, name = "Children on reminder/recall") |>
  rename(Region = region, Village = county)

reminder_by_village_total <- analytic_deid |>
  filter(onreminder == 1) |>
  summarise(`Children on reminder/recall` = n_distinct(patient_id)) |>
  mutate(Region = "CNMI", Village = "CNMI Total", .before = 1)

reminder_by_village <- bind_rows(reminder_by_village_total, reminder_by_village_detail) |>
  arrange(factor(Region, levels = region_levels), Village)

vaccine_needs_detail <- analytic_deid |>
  select(region, county, all_of(need_columns)) |>
  pivot_longer(all_of(need_columns), names_to = "need_variable", values_to = "needed") |>
  mutate(
    Vaccine = names(CNMI_VACCINE_NEED_MAP)[match(need_variable, CNMI_VACCINE_NEED_MAP)]
  ) |>
  group_by(region, county, Vaccine) |>
  summarise(`Children due` = sum(needed == 1, na.rm = TRUE), .groups = "drop") |>
  rename(Region = region, Village = county)

vaccine_needs_total <- analytic_deid |>
  select(all_of(need_columns)) |>
  pivot_longer(all_of(need_columns), names_to = "need_variable", values_to = "needed") |>
  mutate(Vaccine = names(CNMI_VACCINE_NEED_MAP)[match(need_variable, CNMI_VACCINE_NEED_MAP)]) |>
  group_by(Vaccine) |>
  summarise(`Children due` = sum(needed == 1, na.rm = TRUE), .groups = "drop") |>
  mutate(Region = "CNMI", Village = "CNMI Total", .before = 1)

vaccine_needs_table <- bind_rows(vaccine_needs_total, vaccine_needs_detail) |>
  arrange(factor(Region, levels = region_levels), Village, Vaccine)

product_needs_detail <- analytic_deid |>
  select(region, county, all_of(product_columns)) |>
  pivot_longer(all_of(product_columns), names_to = "product_variable", values_to = "needed") |>
  mutate(
    Product = names(CNMI_PRODUCT_MAP)[match(product_variable, CNMI_PRODUCT_MAP)]
  ) |>
  group_by(region, county, Product) |>
  summarise(`Doses needed` = sum(needed == 1, na.rm = TRUE), .groups = "drop") |>
  rename(Region = region, Village = county)

product_needs_total <- analytic_deid |>
  select(all_of(product_columns)) |>
  pivot_longer(all_of(product_columns), names_to = "product_variable", values_to = "needed") |>
  mutate(Product = names(CNMI_PRODUCT_MAP)[match(product_variable, CNMI_PRODUCT_MAP)]) |>
  group_by(Product) |>
  summarise(`Doses needed` = sum(needed == 1, na.rm = TRUE), .groups = "drop") |>
  mutate(Region = "CNMI", Village = "CNMI Total", .before = 1)

product_needs_table <- bind_rows(product_needs_total, product_needs_detail) |>
  arrange(factor(Region, levels = region_levels), Village, Product)

dose_coverage_detail <- analytic_deid |>
  select(region, age_group_label, all_of(unname(CNMI_COVERAGE_MAP))) |>
  pivot_longer(
    all_of(unname(CNMI_COVERAGE_MAP)),
    names_to = "coverage_variable",
    values_to = "utd"
  ) |>
  mutate(
    Vaccine = names(CNMI_COVERAGE_MAP)[match(coverage_variable, CNMI_COVERAGE_MAP)]
  ) |>
  group_by(region, age_group_label, Vaccine) |>
  summarise(
    Denominator = sum(!is.na(utd)),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = if_else(Denominator > 0, 100 * `Children UTD` / Denominator, NA_real_),
    .groups = "drop"
  ) |>
  rename(Region = region, `Age group` = age_group_label)

dose_coverage_total <- analytic_deid |>
  select(age_group_label, all_of(unname(CNMI_COVERAGE_MAP))) |>
  pivot_longer(
    all_of(unname(CNMI_COVERAGE_MAP)),
    names_to = "coverage_variable",
    values_to = "utd"
  ) |>
  mutate(Vaccine = names(CNMI_COVERAGE_MAP)[match(coverage_variable, CNMI_COVERAGE_MAP)]) |>
  group_by(age_group_label, Vaccine) |>
  summarise(
    Denominator = sum(!is.na(utd)),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = if_else(Denominator > 0, 100 * `Children UTD` / Denominator, NA_real_),
    .groups = "drop"
  ) |>
  mutate(Region = "CNMI", .before = 1) |>
  rename(`Age group` = age_group_label)

dose_coverage_table <- bind_rows(dose_coverage_total, dose_coverage_detail) |>
  filter(Denominator > 0) |>
  mutate(
    vaccine_order = match(Vaccine, names(CNMI_COVERAGE_MAP)),
    Vaccine = dose_coverage_label(`Age group`, Vaccine)
  ) |>
  filter(!is.na(Vaccine)) |>
  arrange(
    factor(Region, levels = region_levels),
    factor(`Age group`, levels = age_levels),
    vaccine_order
  ) |>
  select(-vaccine_order)

utd_by_village_detail <- analytic_deid |>
  group_by(region, county, age_group_label) |>
  summarise(
    Denominator = n_distinct(patient_id),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD for age` = 100 * mean(utd == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(Region = region, Village = county, `Age group` = age_group_label)

utd_by_village_total <- analytic_deid |>
  group_by(age_group_label) |>
  summarise(
    Denominator = n_distinct(patient_id),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD for age` = 100 * mean(utd == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(Region = "CNMI", Village = "CNMI Total", .before = 1) |>
  rename(`Age group` = age_group_label)

utd_by_village <- bind_rows(utd_by_village_total, utd_by_village_detail) |>
  arrange(factor(Region, levels = region_levels), Village, factor(`Age group`, levels = age_levels))

coverage_19_35_detail <- analytic_deid |>
  filter(between(age_months, 19, 35)) |>
  select(region, all_of(unname(CNMI_COVERAGE_MAP))) |>
  pivot_longer(
    all_of(unname(CNMI_COVERAGE_MAP)),
    names_to = "coverage_variable",
    values_to = "utd"
  ) |>
  mutate(Vaccine = names(CNMI_COVERAGE_MAP)[match(coverage_variable, CNMI_COVERAGE_MAP)]) |>
  group_by(region, Vaccine) |>
  summarise(
    Denominator = sum(!is.na(utd)),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = if_else(Denominator > 0, 100 * `Children UTD` / Denominator, NA_real_),
    .groups = "drop"
  ) |>
  rename(Region = region)

coverage_19_35_total <- analytic_deid |>
  filter(between(age_months, 19, 35)) |>
  select(all_of(unname(CNMI_COVERAGE_MAP))) |>
  pivot_longer(
    everything(),
    names_to = "coverage_variable",
    values_to = "utd"
  ) |>
  mutate(Vaccine = names(CNMI_COVERAGE_MAP)[match(coverage_variable, CNMI_COVERAGE_MAP)]) |>
  group_by(Vaccine) |>
  summarise(
    Denominator = sum(!is.na(utd)),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = if_else(Denominator > 0, 100 * `Children UTD` / Denominator, NA_real_),
    .groups = "drop"
  ) |>
  mutate(Region = "CNMI", .before = 1)

coverage_19_35 <- bind_rows(coverage_19_35_total, coverage_19_35_detail) |>
  arrange(factor(Region, levels = region_levels), Vaccine)

series_19_35_patient <- analytic_deid |>
  filter(between(age_months, 19, 35)) |>
  mutate(
    `4:3:1:3:3:4 series` = as.integer(
      dtap4utd == 1 & ipv3utd == 1 & mmr1utd == 1 &
        hib3utd == 1 & hepb3utd == 1 & pcv4utd == 1
    ),
    `Routine series + Varicella + HepA` = as.integer(
      `4:3:1:3:3:4 series` == 1 & var1utd == 1 & hepa_utd == 1
    )
  )

series_19_35_detail <- series_19_35_patient |>
  select(region, `4:3:1:3:3:4 series`, `Routine series + Varicella + HepA`) |>
  pivot_longer(-region, names_to = "Series", values_to = "utd") |>
  group_by(region, Series) |>
  summarise(
    Denominator = n(),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = 100 * mean(utd == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(Region = region)

series_19_35_total <- series_19_35_patient |>
  select(`4:3:1:3:3:4 series`, `Routine series + Varicella + HepA`) |>
  pivot_longer(everything(), names_to = "Series", values_to = "utd") |>
  group_by(Series) |>
  summarise(
    Denominator = n(),
    `Children UTD` = sum(utd == 1, na.rm = TRUE),
    `% UTD` = 100 * mean(utd == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(Region = "CNMI", .before = 1)

series_19_35 <- bind_rows(series_19_35_total, series_19_35_detail) |>
  arrange(factor(Region, levels = region_levels), Series)

table_workbook_file <- file.path(
  output_dir,
  paste0("CNMI_Child_Coverage_Planning_Tables_", date_stamp, ".xlsx")
)

title_style <- createStyle(
  fontSize = 13,
  textDecoration = "bold",
  fontColour = "#FFFFFF",
  fgFill = "#0056A6",
  halign = "left",
  valign = "center",
  wrapText = TRUE
)

header_style <- createStyle(
  textDecoration = "bold",
  fontColour = "#FFFFFF",
  fgFill = "#003E73",
  halign = "center",
  valign = "center",
  wrapText = TRUE,
  border = "TopBottomLeftRight",
  borderColour = "#FFFFFF"
)

footnote_style <- createStyle(fontSize = 9, fontColour = "#495057", wrapText = TRUE)
percent_style <- createStyle(numFmt = "0.0", halign = "center")
integer_style <- createStyle(numFmt = "#,##0", halign = "center")

write_table_sheet <- function(wb, sheet, title, data, footnotes = character()) {
  addWorksheet(wb, sheet, gridLines = FALSE)
  n_cols <- max(ncol(data), 1)
  n_rows <- nrow(data)
  data_start <- 4

  writeData(wb, sheet, title, startRow = 1, startCol = 1, colNames = FALSE)
  mergeCells(wb, sheet, cols = 1:n_cols, rows = 1)
  addStyle(wb, sheet, title_style, rows = 1, cols = 1:n_cols, gridExpand = TRUE)
  setRowHeights(wb, sheet, rows = 1, heights = 34)

  writeData(wb, sheet, data, startRow = 3, startCol = 1, keepNA = FALSE)
  addStyle(wb, sheet, header_style, rows = 3, cols = 1:n_cols, gridExpand = TRUE)
  addFilter(wb, sheet, rows = 3, cols = 1:n_cols)

  if (n_rows > 0) {
    body_rows <- data_start:(data_start + n_rows - 1)
    percent_cols <- which(str_detect(names(data), "^%"))
    integer_cols <- which(str_detect(names(data), "Children|Denominator|Doses"))

    if (length(percent_cols) > 0) {
      addStyle(wb, sheet, percent_style, rows = body_rows, cols = percent_cols, gridExpand = TRUE)
    }
    if (length(integer_cols) > 0) {
      addStyle(wb, sheet, integer_style, rows = body_rows, cols = integer_cols, gridExpand = TRUE)
    }
  }

  if (length(footnotes) > 0) {
    footnote_start <- data_start + n_rows + 1
    for (i in seq_along(footnotes)) {
      row <- footnote_start + i - 1
      writeData(wb, sheet, footnotes[[i]], startRow = row, startCol = 1, colNames = FALSE)
      mergeCells(wb, sheet, cols = 1:n_cols, rows = row)
      addStyle(wb, sheet, footnote_style, rows = row, cols = 1:n_cols, gridExpand = TRUE)
    }
  }

  freezePane(wb, sheet, firstActiveRow = 4)
  setColWidths(wb, sheet, cols = 1:n_cols, widths = "auto")
}

schedule_footnotes <- c(
  "MMR and varicella dose 2 are not required before age 4 years (48 months).",
  "HepA dose 2 is due at least 6 months after dose 1.",
  "Schedule source: https://downloads.aap.org/AAP/PDF/AAP-Immunization-Schedule.pdf"
)

wb <- createWorkbook()

write_table_sheet(
  wb, "IIS Denominator",
  paste0("Active CNMI IIS patients aged 2-83 months, by region and age group - ", year(analysis_date)),
  denominator_table
)
write_table_sheet(
  wb, "Reminder by Village",
  paste0("Children on reminder/recall, by CNMI region and village - ", year(analysis_date)),
  reminder_by_village
)
write_table_sheet(
  wb, "Needs by Vaccine",
  paste0("Children needing vaccination, by CNMI region, village, and vaccine - ", year(analysis_date)),
  vaccine_needs_table,
  schedule_footnotes
)
write_table_sheet(
  wb, "Products Needed",
  paste0("Estimated vaccine products needed, by CNMI region and village - ", year(analysis_date)),
  product_needs_table,
  "Product estimates are planning counts; clinical review is required before administration."
)
write_table_sheet(
  wb, "Dose Coverage",
  paste0("Age-appropriate vaccination coverage, by CNMI region and age group - ", year(analysis_date)),
  dose_coverage_table,
  c(
    schedule_footnotes,
    paste0(
      "* DTaP is up to date with 5 doses, or 4 when dose 4 was given at age 48 months or older; ",
      "IPV is up to date with 4 doses, or 3 when dose 3 was given at age 48 months or older."
    )
  )
)
write_table_sheet(
  wb, "UTD by Village",
  paste0("Up-to-date status, by CNMI region, village, and age group - ", year(analysis_date)),
  utd_by_village,
  schedule_footnotes
)
write_table_sheet(
  wb, "Coverage 19-35 Months",
  paste0("Vaccination coverage among children aged 19-35 months, by CNMI region - ", year(analysis_date)),
  coverage_19_35,
  schedule_footnotes
)
write_table_sheet(
  wb, "Series 19-35 Months",
  paste0("Vaccination series coverage among children aged 19-35 months, by CNMI region - ", year(analysis_date)),
  series_19_35,
  c(
    "4:3:1:3:3:4 = >=4 DTaP, >=3 IPV, >=1 MMR, >=3 Hib, >=3 HepB, and >=4 PCV.",
    "The expanded routine series adds >=1 varicella dose and age-appropriate HepA status."
  )
)

saveWorkbook(wb, table_workbook_file, overwrite = TRUE)
```

# Run summary

```{r}
#| label: run-summary

run_summary <- tibble(
  Measure = c(
    "Active children aged 2-83 months",
    "Children UTD for age",
    "Children on reminder/recall",
    "Villages without coordinates"
  ),
  Value = c(
    n_distinct(analytic_deid$patient_id),
    sum(analytic_deid$utd == 1, na.rm = TRUE),
    sum(analytic_deid$onreminder == 1, na.rm = TRUE),
    nrow(missing_coordinates)
  )
)

knitr::kable(run_summary)
```

Files created in `data/Analytic Code Output`:

```{r}
#| label: output-files
#| echo: false

tibble(
  Output = c(
    "Full patient-level dataset",
    "Deidentified dashboard dataset",
    "Patients needing vaccination",
    "Formatted planning tables",
    "Missing-coordinate QA"
  ),
  File = basename(c(
    full_dataset_file,
    deid_dataset_file,
    patients_needing_file,
    table_workbook_file,
    missing_coordinates_file
  ))
) |>
  knitr::kable()
```
